---
title: "Final Report"
author: "Shan Sun, Faith Wong"
date: "STAT 335, Fall 2023"
output: pdf_document
fontsize: 12pt
header-includes:
  - \usepackage[margin=1in]{geometry}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

<!-- An introduction to the data and the problem would be a very good way to start. -->

The dataset describes U.S. flight delays from January 2017 - July 2022. Some variables included are the year and month of data collection, airline carrier, airport, the number of flights delayed due to a specific reason (air carrier, weather, National Aviation System, security issue, previous flight delayed), and the total time (minutes) of delay due to a specific reason.

## The Research Question

<!-- Clearly state your SINGLE RESEARCH QUESTION -->
Research Question: **What is the relationship between the both and the number of delays**

## Analysis

<!-- Your analysis goes here.  -->

<!-- Be sure to carefully select the plots and graphics that best support your analysis and conclusion. -->

We began by loading the data and manipulating it into a format we can use. As described in Phase 2, we created a column for dates, removed categorical variables, and aggregated the data for each month. Also, a data frame is created for pre-COVID data, since we previously saw that COVID had a large impact on flights.

```{r, results='hide'}
library(astsa)
delays <- read.csv("https://raw.githubusercontent.com/nekalik/STAT-335-Project/main/Airline_Delay_Cause.csv")
delays$date = as.Date(paste(delays$year, delays$month, 1, sep = "-"))
sub_delays = delays[-(3:6)]
sum_delays= aggregate(.~date, sub_delays,sum)
sum_delays_pre = sum_delays[sum_delays$date<'2020-01-01', ]
```

For now, we focus on the number of delayed flights ("arr_del15") before COVID.

```{r}
#a general name for variable modeling
x = sum_delays_pre$arr_del15

tsplot(x)
lm.x = lm(x~sum_delays_pre$date)

plot(x=sum_delays_pre$date, y=x)
abline(lm.x)

summary(lm.x)
par(mfrow = c(2, 2))
plot(lm.x)
```

The time series plot of x shows a linear trend. After performing a linear regression on the model on time, there seems to be some nonconstant variance in the residual plot, so to help with model selection, we also built some models with log(x)



<!-- ## Model Building With X -->
<!-- ```{R} -->
<!-- tsplot(x) -->

<!-- #acf within the range after the first few lag -->
<!-- acf(x,30) -->
<!-- acf2(x,30) -->
<!-- ``` -->
<!-- AIC cut off after lag 2, and PACF cut off after lag 1 -->

<!-- ```{R} -->
<!-- sarima(x, p=1, d=0, q =0) -->
<!-- sarima(x, p=0, d=0, q=2) -->
<!-- ``` -->
<!-- For both models, the variable are significant, and the residual looks like a white noise and pass the Ljung-Box test -->

<!-- Also tried building model for by detrending it -->

<!-- ```{R} -->
<!-- #detrend -->
<!-- tsplot(diff(x)) -->
<!-- #more stationary looking -->
<!-- acf(diff(x),30) -->
<!-- #parameter selection -->
<!-- acf2(diff(x),30) -->
<!-- sarima(x, p=0, d=1, q=2) -->
<!-- ``` -->
<!-- Potentially a ARIMA(0,1,2) model. The AIC for the past 3 models are very close.  -->


<!-- ## Model Building with log(X) -->
<!-- ```{R} -->
<!-- par(mfrow = c(2, 1)) -->
<!-- tsplot(log(x)) -->
<!-- tsplot(diff(log(x))) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- acf2(diff(log(x)), 20) -->
<!-- ``` -->
<!-- The ACF of the model seems to be touching the dotted line at lag of 1, so we tried both p=0, and p=1. -->
<!-- The PACF of the model seems to cut off after lag of 1.  -->

<!-- ```{r} -->
<!-- sarima(log(x), p=1, d=1, q =0, no.constant = TRUE) -->
<!-- sarima(log(x), p=1, d=1, q =1, no.constant = TRUE) -->
<!-- ``` -->
<!-- The estimate of ma1 is not significant and the AIC of ARIMA(1,1,0) is little bit lower than AIC of ARIMA(1,1,1), so ARIMA(1,1,0) is the better out of the two.  -->

<!-- ## Model Selection with Seasonality -->
<!-- ```{r} -->
<!-- tsplot(diff(diff(x),12)) -->
<!-- acf2(diff(diff(x),12), 20) -->


<!-- #sarima(log(x), p=0, d=1, q=0, P=0, D=1, Q=0, S=12) -->
<!-- ``` -->
<!-- The acf and pacf goes to 0 after lag of 0, so it there doesn't seems to be seaonality within the data -->

```{r, results='hide'}
sar.x = sarima(log(x), p=1, d=1, q =1, no.constant = TRUE)
sar.x
```








## Conclusion

<!-- Be sure to provide a clearly articulated answer to your research question that is well-supported by your analysis. -->

```{r, results='hide'}
sarima(log(x), p=1, d=1, q =1, no.constant = TRUE)

par(mfrow=c(2,1))

#still need to work on this!
sarima.for(log(x), p=1, d=1, q =1, no.constant = TRUE, n.ahead=24)

tsplot(log(sum_delays$arr_del15), xlim=c(0,60), ylim=c(9,12.5))
```

Comparing the prediction of delays if covid didn't hit.


when the delays is back to normal (or when covid is offically over based on delay counts)
## References

<!-- I'm not picky about reference style but references should be accurate and complete. If you want a style guide, use this one: [IEEE Style Guide](https://ieee-dataport.org/sites/default/files/analysis/27/IEEE%20Citation%20Guidelines.pdf) -->



The data was found on Kaggle.com:

Jawad Khattak, 2022. *"Flight Delay from January 2017 - July 2022."* Kaggle.com, https://www.kaggle.com/datasets/jawadkhattak/us-flight-delay-from-january-2017-july-2022.

The original source of the Kaggle.com data:

U.S. Department of Transportation: Bureau of Transportation Statistics (BTS), 2022. *"Airline On-Time Statistics and Delay Causes."* BTS website, https://www.transtats.bts.gov/OT_Delay/OT_DelayCause1.asp.
