---
title: "STAT 335 Project Phase 2"
author: "Shan Sun, Faith Wong"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

```{r}
# loading time series packages
library(xts)
library(astsa)
```

```{r}
# reading in csv file to R as "delays" dataframe
delays <- read.csv("https://raw.githubusercontent.com/nekalik/STAT-335-Project/main/Airline_Delay_Cause.csv")
```

To restate from the Phase 1 submission:

- Data description: Flight delays from January 2017 - July 2022

- Research question: What is the relationship between the month and the number and nature of delays?

# Data Cleaning/Organization

## Add dates

The data had one column for the year and one column for the month, so we created a new column that combined them.

```{R}
delays$date = as.Date(paste(delays$year, delays$month, 1, sep = "-"))
head(delays)
```

# Step 1: Overall Delays

We began our exploratory data analysis by looking at overall delays over the dataset's entire time frame. 

## Aggregation

Since each date had multiple entries, we needed to aggregate the data so there would be only one row of values for each month. 

To do this, we first eliminate categorical variables such as carrier and airport (hard to perform aggregation on these).

```{R}
sub_delays = delays[-(3:6)]
head(sub_delays)
```

Next, we aggregated the data based on each column's sum.

```{R}
# finding sum of each column by date
sum_delays= aggregate(.~date, sub_delays,sum)
head(sum_delays)
```

## Data exploration

We played around with the aggregated data and tried to get a better idea of delays as a whole. To measure delays, we looked at both the number of delays as well as how much planes were delayed.

```{r}
#plot the total number of delayed flights over time
tsplot(sum_delays$date, sum_delays$arr_del15)

#plot the total delay in mins over time
tsplot(sum_delays$date, sum_delays$arr_delay)
```

The two time series plots look approximately the same visually, which makes sense because the total delay in minutes is dependent on how many delays there are. Thus, we can analyze either measure to understand flight delays over time.

We also check to see if we can apply a classical structural model to the data.

This code was used:

```{r, eval=FALSE}
x = window(sum_delays$arr_del15)
plot(stl(x, s.window="per"))
```

However, R gave an error saying "series is not periodic or has less than two periods" - we cannot use this model.

Let's try fitting a simple linear regression model.

```{r}
#basic regression model - delay in minutes
lm = lm(arr_delay~date, data = sum_delays)
summary(lm)
```

As we can see from the R-squared value, the regression model doesn't fit the data very well, likely due to the big dip seen in the time series plots starting around 2020. It is reasonable to assume that COVID-19 has drastically impacted air travel in the U.S. 

Since it is difficult to "remove" COVID's impact after 2020, we decide to focus on pre-COVID data, narrowing our data to span January 2017-December 2019.

# Step 2: Pre-COVID Delays

We again do a general EDA on pre-COVID data.

## Subsetting dataset

```{R}
# isolating data to pre-Covid only (January 2017-December 2019)
sum_delays_pre = sum_delays[sum_delays$date<'2020-01-01', ]
```

## Exploring data

We first view our "new" time series.

```{r}
#plot the total delay in mins over time
ts.plot(sum_delays_pre$arr_delay)
```

We see a possible slight upward trend... let's check that out.

```{R}
#fitting SLR model
trend = lm(arr_delay~date, data = sum_delays_pre)
summary(trend)

#plot data and trend
tsplot(sum_delays_pre$date, sum_delays_pre$arr_delay)
abline(trend,col="red")

#plot residual
par(mfrow=c(2,2)) 
plot(trend)
```

The p-values for the model and coefficients are significant - a relationship exists between time and the total delay time (delays increase over time). 

Looking at the residuals, we see that our pre-COVID data approximately meets the regression assumptions. Although the Residuals vs Fitted line isn't perfectly straight, it does center around 0, and the residuals are pretty randomly scattered. 

We wonder if the data is heteroscedastic:

```{R}
#taking log of data to address possible non-constant variance
test = lm(log(arr_delay)~date, data = sum_delays_pre)
summary(test)

tsplot(sum_delays_pre$date, log(sum_delays_pre$arr_delay))
abline(test,col="red")

par(mfrow=c(2,2)) 
plot(test)
```

Doing a log transformation does not help with modeling the data, so non-constant variance isn't an issue.

Let's investigate the data some more using lagplots.

```{R}
# lagplot for number of delays, for 12 months
lag1.plot(sum_delays_pre$arr_del15,12)
```
Lag=1 has the strongest correlation, meaning that the current data is most strongly related to the data 1 month before. The correlation is 0.57, meaning the current month's delays is somewhat dependent on the previous month's. 

Based on our analyses on the pre-COVID data so far, it seems like overall delays may be some sort of autoregressive or random walk model.

Let's return to the linear model and see what happens when we remove the trend (as we see that there's large variation around the trend line - maybe something else is going on).

```{R}
#detrending
tsplot(resid(trend))
acf1(resid(trend),30)
```

```{R}
#differencing
tsplot(diff(sum_delays_pre$arr_delay))
acf1(diff(sum_delays_pre$arr_delay),30)
```

With the trend removed, the variation is relatively random and indicative of white noise. Most importantly, there seems to be seasonality corresponding to at least the year (12 months) and possibly even some for approx. every half year.

Detrending seems to highlight the annual cycle, while differencing shows some possible smaller cycles. Information can be gained by both of these methods, but the conclusion from detrending seems to be the most obvious.
