---
title: "STAT 335 Project Phase 2"
author: "Shan Sun, Faith Wong"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

```{r}
# loading time series packages
library(xts)
library(astsa)
```

```{r}
# reading in csv file to R as "delays" dataframe
delays <- read.csv("https://raw.githubusercontent.com/nekalik/STAT-335-Project/main/Airline_Delay_Cause.csv")
```

To restate from the Phase 1 submission:

- Data description: Flight delays from January 2017 - July 2022

- Research question: What is the relationship between the month and the number and nature of delays?

# Data Cleaning/Organization

## Add dates

The data had one column for the year and one column for the month, so we created a new column that combined them.

```{R}
delays$date = as.Date(paste(delays$year, delays$month, 1, sep = "-"))
head(delays)
```

# Step 1: Overall Delays

We began our exploratory data analysis by looking at overall delays over the dataset's entire time frame. 

## Aggregation

Since each date had multiple entries, we needed to aggregate the data so there would be only one row of values for each month. 

To do this, we first eliminate categorical variables such as carrier and airport (hard to perform aggregation on these).

```{R}
sub_delays = delays[-(3:6)]
head(sub_delays)
```

Next, we aggregated the data based on each column's sum.

```{R}
# finding sum of each column by date
sum_delays= aggregate(.~date, sub_delays,sum)
head(sum_delays)
```

## Data exploration

We played around with the aggregated data and tried to get a better idea of delays as a whole. To measure delays, we looked at both the number of delays as well as how much planes were delayed.

```{r}
#plot the total number of delayed flights over time
tsplot(sum_delays$date, sum_delays$arr_del15)

#plot the total delay in mins over time
tsplot(sum_delays$date, sum_delays$arr_delay)
```

The two time series plots look approximately the same visually, which makes sense because the total delay in minutes is dependent on how many delays there are. Thus, we can analyze either measure to understand flight delays over time.

```{r}
#ACF of delay in minutes
acf1(sum_delays$arr_delay, max.lag = 48)
```

The ACF isn't strongly indicative of a certain time series model... perhaps the dip from 2020-2021 is impacting it. We will elaborate below.

```{r}
#basic regression model - delay in minutes
lm = lm(arr_delay~date, data = sum_delays)
summary(lm)
```

As we can see from the R-squared value, the regression model doesn't fit the data very well, likely due to the big dip seen in the time series plots starting around 2020. It is reasonable to assume that COVID-19 has drastically impacted air travel in the U.S. 

Since it is difficult to "remove" COVID's impact after 2020, we decide to focus on pre-COVID data, narrowing our data to span January 2017-December 2019.

# Step 2: Pre-COVID Delays

We again do a general EDA on pre-COVID data.
