---
title: "STAT 335 Project Phase 2"
author: "Shan Sun, Faith Wong"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Phase 1

```{r, include=false}
# loading time series packages
library(xts)
library(astsa)
```

## Data

```{r}
# reading in csv file to R as "delays" dataframe
delays <- read.csv("/Users/Shan/Downloads/Airline_Delay_Cause.csv")
```

```{r}
# first look at data
str(delays)
head(delays)
```
## Data description

Jawad Khattak, 2022. *"Flight Delay from January 2017 - July 2022."* Kaggle.com, https://www.kaggle.com/datasets/jawadkhattak/us-flight-delay-from-january-2017-july-2022


## Research question

**What is the relationship between the month and the number and nature of delays?**

For example:

Do the primary causes of delays vary from month to month? Is there a dominant reason that causes the most delays? Are certain months associated with a certain kind of delay? How does the month impact how many total delays there are?

# Phase 2

Add Dates
```{R}
delays$date = as.Date(paste(delays$year, delays$month, 1, sep = "-"))
head(delays)
```

Eliminate categorical variable 
```{R}
sub_delays = delays[-(3:6)]
head(sub_delays)
```

## Sum delay for each month
```{R}
# finding sum of each column by date
sum_delays= aggregate(.~date, sub_delays,sum)
head(sum_delays)

#plot the trend of total number of flights that is more than 15 min late 
#look rough the same as plotting arr_delay, so for now on the doc, we use arr_delay (why this over other?)
tsplot(sum_delays$date, sum_delays$arr_del15)

```
SLR 
```{R}
#plot, total delay in mins 
tsplot(sum_delays$date, sum_delays$arr_delay)

#basic regression model
lm = lm(arr_delay~date, data = sum_delays)
summary(lm)

#ACF of SLR
acf1(sum_delays$arr_delay, max.lag = 48)

```

## Mean delay for each month
```{R}

mean_delays= aggregate(.~date, sub_delays,mean)

```

SLR

```{R}
#plot
tsplot(mean_delays$date, mean_delays$arr_delay)

#basic regression model
lm = lm(arr_delay~date, data = mean_delays)
summary(lm)

#ACF of SLR
acf1(mean_delays$arr_delay, max.lag = 48)

```
## Focus on pre-covid
```{R}
# isolating data to pre-Covid only (January 2017-December 2019)
sum_delays_pre = sum_delays[sum_delays$date<'2020-01-01', ]

mean_delays_pre = mean_delays[mean_delays$date<'2020-01-01', ]

```

Plot
```{R}
#different trends here, wonder why 
par(mfrow = c(1,2))
ts.plot(sum_delays_pre$arr_delay)
ts.plot(mean_delays_pre$arr_delay)

```
```{R}
ts.plot(sum_delays_pre$carrier_ct/sum_delays_pre$arr_flights, col = "red", ylim = c(0,.10))
lines(sum_delays_pre$weather_ct/sum_delays_pre$arr_flights, col="green")
lines(sum_delays_pre$nas_ct/sum_delays_pre$arr_flights, col="blue")
lines(sum_delays_pre$security_ct/sum_delays_pre$arr_flights, col="purple")
lines(sum_delays_pre$late_aircraft_ct/sum_delays_pre$arr_flights, col="black")

acf(sum_delays_pre$late_aircraft_ct/sum_delays_pre$arr_flights, 36)
```

<!-- Plot -->
<!-- ```{R} -->
<!-- #different trends here, wonder why  -->
<!-- par(mfrow = c(1,2)) -->
<!-- ts.plot(sum_delays_pre$arr_del15) -->
<!-- ts.plot(mean_delays_pre$arr_del15) -->

<!-- ``` -->

SLR 
```{R}
#SLR of pre-covid data
trend = lm(arr_delay~date, data = sum_delays_pre)
summary(trend)

#plot data and trend
tsplot(sum_delays_pre$date, sum_delays_pre$arr_delay)
abline(trend,col="red")

#plot residual
par(mfrow=c(2,2)) 
plot(trend)
```
SLR 
```{R}
#SLR of pre-covid data
trend = lm(arr_delay~date, data = mean_delays_pre)
summary(trend)

#plot data and trend
tsplot(mean_delays_pre$date, mean_delays_pre$arr_delay)
abline(trend,col="red")

#plot residual
par(mfrow=c(2,2)) 
plot(trend)

#Check hetero and does not help
test = lm(log(arr_delay)~date, data = sum_delays_pre)
summary(test)

tsplot(sum_delays_pre$date, log(sum_delays_pre$arr_delay))
abline(test,col="red")

par(mfrow=c(2,2)) 
plot(test)
```

```{R}

lag1.plot(sum_delays_pre$arr_del15,12)

```

##ratio
```{R}
sum_delays_pre["ratio"] = sum_delays_pre$arr_del15/sum_delays_pre$arr_flights
ts.plot(sum_delays_pre$ratio)

acf(sum_delays_pre$ratio)

x = as.data.frame(sum_delays_pre)
head(x)
#plot(decompose(x$ratio))

```

#look at weather
```{R}
sum_delays_pre["ratio"] = sum_delays_pre$weather_ct/sum_delays_pre$arr_flights
ts.plot(sum_delays_pre$ratio)

acf(sum_delays_pre$ratio, 24)

x = as.data.frame(sum_delays_pre)
head(x)

```



## carrier
```{R}

carrier_unprocessed = cbind("carrier" = delays$carrier,sub_delays) 
head(carrier_unprocessed)
carrier= aggregate(.~date+carrier, carrier_unprocessed, mean)

carrier_pre = carrier[carrier$date<'2020-01-01', ]

carrier_pre["ratio"] = carrier_pre$arr_del15/carrier_pre$arr_flights

head(carrier_pre,20)

```

```{R}
diff_carrier = unique(carrier_pre$carrier)
ts.plot(carrier_pre$ratio[carrier_pre$carrier=="9E"], ylim= c(0,.5), ylab ="delays")
num = 1
for(x in diff_carrier){
  lines(carrier_pre$ratio[carrier_pre$carrier==x], col=num, lwd = 2 )
   num = num +1

     }
```

```{R}

par(mfrow=c(3,6))
for(x in diff_carrier){ 
  ts.plot(carrier_pre$ratio[carrier_pre$carrier==x], ylab="delays", ylim= c(0,.5)) 
  }
```
```{R}
ts.plot(carrier_pre$ratio[carrier_pre$carrier=="AA"],xlab = "9E", ylab="delays", ylim= c(0,.5)) 
lines(carrier_pre$ratio[carrier_pre$carrier=="B6"], col = "blue")
lines(carrier_pre$ratio[carrier_pre$carrier=="EV"], col = "red") 
lines(carrier_pre$ratio[carrier_pre$carrier=="F9"], col = "purple")
lines(carrier_pre$ratio[carrier_pre$carrier=="VX"], col = "orange")
lines(carrier_pre$ratio[carrier_pre$carrier=="G4"], col = "pink") 
```

```{R} 
ts.plot(carrier_pre$ratio[carrier_pre$carrier=="9E"],xlab = "9E", ylab="delays", ylim= c(0,.5), col="dark green", xlim = c(0,35)) 
lines(carrier_pre$ratio[carrier_pre$carrier=="AS"], col = "blue")
lines(carrier_pre$ratio[carrier_pre$carrier=="DL"], col = "red") 
lines(carrier_pre$ratio[carrier_pre$carrier=="HA"], col = "purple")
lines(carrier_pre$ratio[carrier_pre$carrier=="OO"], col = "green")
lines(carrier_pre$ratio[carrier_pre$carrier=="WN"], col = "orange")
lines(carrier_pre$ratio[carrier_pre$carrier=="YX"], col = "pink") 
```
## regression model

```{r}
#remove ones with less data
carrier_full = data.frame()
for(a in diff_carrier){ 
  x = carrier_pre$arr_flights[carrier_pre$carrier==a]
  y = carrier_pre$ratio[carrier_pre$carrier==a]
  if(length(x)==36){
    carrier_full = rbind(carrier_full, carrier_pre[which(carrier_pre$carrier==a),])
  }
}

```

```{R}
#plot ratio of delay vs time for each carrier in separate plot
diff_carrier = unique(carrier_full$carrier)

par(mfrow=c(3,6))
for(x in diff_carrier){ 
  ts.plot(carrier_full$ratio[carrier_full$carrier==x], ylab="delays", ylim= c(0,.4)) 
  }

```

```{r}
#plot ratio of delay vs time for each carrier on the same plot
ts.plot(carrier_full$ratio[carrier_full$carrier=="AA"], ylim= c(0,.4), ylab ="delays")
num = 1
for(x in diff_carrier){
  lines(carrier_full$ratio[carrier_full$carrier==x], col=num, lwd = 1 )
   num = num +1

     }
```


```{r}
#plot the ratio of delay as a function of number of flights for each carrier 
par(mfrow=c(3,6))
for(x in diff_carrier){ 
  plot(carrier_full$arr_flights[carrier_full$carrier==x],carrier_full$carrier_ct[carrier_full$carrier==x], xlab = x, ylab = "delay ratio")
}
```

```{R}
#build a linear model for each carrier (carrier_ct ~ arr_flight)
for(a in diff_carrier){ 
  x = carrier_full$arr_flights[carrier_full$carrier==a]
  y = carrier_full$carrier_ct[carrier_full$carrier==a]
  cor(x,y)
  lm.x = lm(y~x)
  print(summary(lm.x))
}
```



Only got through here


```{R}
library(corrplot)
corrplot(cor(sum_delays_pre[-(1:3)]))
```

```{R}
#detrending
tsplot(resid(trend))
acf1(resid(trend),30)


```

```{R}

#differencing
tsplot(diff(sum_delays_pre$arr_delay))
acf1(diff(sum_delays_pre$arr_delay),30)

```


```{R}

summary_delays[19] = c(1:36)

smooth = lowess(summary_delays$arr_del15, f=.4)

tsplot(smooth)
summary_delays[20] = smooth[2]
head(summary_delays)
trend = lm(x~V19 + arr_flights, data = summary_delays)
summary(trend)
tsplot(summary_delays$date, summary_delays$x)
lines(63.44,-0.56,0.0510,col="red")
par(mfrow=c(2,2)) 
plot(trend)
```

