---
title: "STAT 335 Project Phase 2"
author: "Shan Sun, Faith Wong"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Phase 1

```{r, include=false}
# loading time series packages
library(xts)
library(astsa)
```

## Data

```{r}
# reading in csv file to R as "delays" dataframe
delays <- read.csv("https://raw.githubusercontent.com/nekalik/STAT-335-Project/main/Airline_Delay_Cause.csv")
```

```{r}
# first look at data
str(delays)
head(delays)
```
## Data description

Jawad Khattak, 2022. *"Flight Delay from January 2017 - July 2022."* Kaggle.com, https://www.kaggle.com/datasets/jawadkhattak/us-flight-delay-from-january-2017-july-2022


## Research question

**What is the relationship between the month and the number and nature of delays?**

For example:

Do the primary causes of delays vary from month to month? Is there a dominant reason that causes the most delays? Are certain months associated with a certain kind of delay? How does the month impact how many total delays there are?

# Phase 2

Add Dates
```{R}
delays$date = as.Date(paste(delays$year, delays$month, 1, sep = "-"))
head(delays)
```

Eliminate categorical variable 
```{R}
sub_delays = delays[-(3:6)]
head(sub_delays)
```

## Sum delay for each month
```{R}
# finding sum of each column by date
sum_delays= aggregate(.~date, sub_delays,sum)
head(sum_delays)

#plot the trend of total number of flights that is more than 15 min late 
#look rough the same as plotting arr_delay, so for now on the doc, we use arr_delay (why this over other?)
tsplot(sum_delays$date, sum_delays$arr_del15)

```
SLR 
```{R}
#plot, total delay in mins 
tsplot(sum_delays$date, sum_delays$arr_delay)

#basic regression model
lm = lm(arr_delay~date, data = sum_delays)
summary(lm)

#ACF of SLR
acf1(sum_delays$arr_delay, max.lag = 48)

```

## Mean delay for each month
```{R}

mean_delays= aggregate(.~date, sub_delays,mean)

```

SLR

```{R}
#plot
tsplot(mean_delays$date, mean_delays$arr_delay)

#basic regression model
lm = lm(arr_delay~date, data = mean_delays)
summary(lm)

#ACF of SLR
acf1(mean_delays$arr_delay, max.lag = 48)

```
## Focus on pre-covid
```{R}
# isolating data to pre-Covid only (January 2017-December 2019)
sum_delays_pre = sum_delays[sum_delays$date<'2020-01-01', ]

mean_delays_pre = mean_delays[mean_delays$date<'2020-01-01', ]

```

Plot
```{R}
#different trends here, wonder why 
par(mfrow = c(1,2))
ts.plot(sum_delays_pre$arr_delay)
ts.plot(mean_delays_pre$arr_delay)

```


<!-- Plot -->
<!-- ```{R} -->
<!-- #different trends here, wonder why  -->
<!-- par(mfrow = c(1,2)) -->
<!-- ts.plot(sum_delays_pre$arr_del15) -->
<!-- ts.plot(mean_delays_pre$arr_del15) -->

<!-- ``` -->

SLR 
```{R}
#SLR of pre-covid data
trend = lm(arr_delay~date, data = sum_delays_pre)
summary(trend)

#plot data and trend
tsplot(sum_delays_pre$date, sum_delays_pre$arr_delay)
abline(trend,col="red")

#plot residual
par(mfrow=c(2,2)) 
plot(trend)
```
SLR 
```{R}
#SLR of pre-covid data
trend = lm(arr_delay~date, data = mean_delays_pre)
summary(trend)

#plot data and trend
tsplot(mean_delays_pre$date, mean_delays_pre$arr_delay)
abline(trend,col="red")

#plot residual
par(mfrow=c(2,2)) 
plot(trend)
```

Only got through here


```{R}
#Check hetero and does not help
test = lm(log(arr_flights)~date, data = summary_delays)
summary(test)

tsplot(summary_delays$date, log(summary_delays$arr_flights))
abline(test,col="red")

par(mfrow=c(2,2)) 
plot(test)
```

```{R}
library(corrplot)
corrplot(cor(summary_delays[-(1:3)]))
```

```{R}
#detrending
tsplot(resid(trend))
acf1(resid(trend),30)


```

```{R}

#differencing
tsplot(diff(summary_delays$arr_flights))
acf1(diff(summary_delays$arr_flights),30)

```


```{R}

summary_delays[19] = c(1:36)

smooth = lowess(summary_delays$arr_del15, f=.4)

tsplot(smooth)
summary_delays[20] = smooth[2]
head(summary_delays)
trend = lm(x~V19 + arr_flights, data = summary_delays)
summary(trend)
tsplot(summary_delays$date, summary_delays$x)
lines(63.44,-0.56,0.0510,col="red")
par(mfrow=c(2,2)) 
plot(trend)
```



